Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-25 manuscript.rmd
1. added references to Brum et al.

Our study brings attention to the need for assessing drought sensitivity in terms of species realized drought exposure by accounting for the actual hydrological drought and trees' deep water access, beyond the universal drought indices and trees' above-ground hydraulic vulnerability. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through periods of extreme droughts (Figs. \@ref(fig:erd-mort-census-evg), S\@ref(fig:erd-mort-census), \@ref(fig:swp-droughts)). Deeper ERD species on average had more hydraulically efficient, but risky, suite of traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk. By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{soil,~z = ERD}$ crossed ERD $\mathrm{\Psi}_{crit}$. Our result of deep rooted species being more hydraulically vulnerble, is consistent with a seasonal forest in the Amazon, where among 12 co-occurring species trees sourcing deeper water were 50 and 70% of variance in $\mathrm{\Psi}_\textrm{88,stem}$ ($\mathrm{\delta}$^18^O~xylem~) Our results are also consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016]. @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\mathrm{\Psi}$~88,stem~) and had narrower safety-margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~). @Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\mathrm{\Psi}$~88,stem~ and narrower safety margin compared to two species sourcing shallow-water.

>>>>

Our study brings attention to the need for assessing drought sensitivity in terms of species realized drought exposure by accounting for the actual hydrological drought and trees' deep water access, beyond the universal drought indices and trees' above-ground hydraulic vulnerability. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through periods of extreme droughts (Figs. \@ref(fig:erd-mort-census-evg), S\@ref(fig:erd-mort-census), \@ref(fig:swp-droughts)). Deeper ERD species on average had more hydraulically efficient, but risky, suite of traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk. By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{soil,~z = ERD}$ crossed ERD $\mathrm{\Psi}_{crit}$. Our result of deep rooted species being more hydraulically vulnerable is consistent with observations for 12 co-occurring species in a seasonal forest in the Amazon, amongst which species $\mathrm{\delta}$^18^O~xylem~ explained 70% variance in species $\mathrm{\Psi}_\textrm{88,stem}$ [@Brum.2019]. Our results are also consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016]. @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\mathrm{\Psi}$~88,stem~) and had narrower safety-margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~). @Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\mathrm{\Psi}$~88,stem~ and narrower safety margin compared to two species sourcing shallow-water.
